Estimating Predictive Rate-Distortion Curves via Neural Variational Inference

### 2019-05-11 paper.tex
minor changes for submission

\newcounter{theorem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{question}[theorem]{Question}
\newtheorem{example}[theorem]{Example}
\newtheorem{defin}[theorem]{Definition}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{thm}[theorem]{Theorem}

@@

% Reviewers:
% Susanne Still
% Lukasz Debowski
% Kumiko Tanaka-Ishii
% Noga Zaslavsky
% Alexander Alemi

% TODO mention sam bowman's language VAE

% potentially cite https://arxiv.org/pdf/1609.05191.pdf

@@

The distortion is the loss in predictive accuracy when predicting from $Z$, relative to optimal prediction from the full past $\past$.
In the Information Bottleneck formalization, this is equivalent to the amount of mutual information between past and future that is \emph{not} captured by $Z$ \citep{harremoes2007information}: %\rljf{Cite the paper "how to choose a good distortion measure"}
\begin{equation}
	\operatorname{I}[\past, \future|Z].
\end{equation}
Due to the Markov condition, the distortion measure satisfies the relation
\begin{equation}
	\operatorname{I}[\past, \future|Z] = \operatorname{I}[\past, \future] - \operatorname{I}[Z, \future],
\end{equation}
i.e., it captures how much less information $Z$ carries about the future $\future$ compared to the full past $\past$.
For a fixed process $(X_t)_t$, choosing $Z$ to minimize the distortion is equivalent to maximizing the mutual information between the code and the future:
\begin{equation}\label{eq:predictiveness}
\operatorname{I}[Z, \future]
\end{equation}
We will refer to~(\ref{eq:predictiveness}) as the \key{predictiveness} of the code $Z$.

@@

As discussed in Section~\ref{sec:complexity}, estimating predictive rate-distortion is related to the problem of estimating statistical complexity.
\citet{clarke2003application} and \citet{still-optimal-2010} consider the problem of estimating statistical complexity from finite data.
While statistical complexity cannot be identified from finite data in general, \citet{clarke2003application} introduce certain restrictions on the underlying process that make this more feasible.

@@

The main prior method for estimating predictive rate-distortion from data is Optimal Causal Filtering (OCF, \citet{still-optimal-2010}).
This method approximates predictive rate-distortion using two approximations:
First, it replaces semi-infinite pasts and futures with bounded-length contexts, i.e., pairs of finite past contexts $(\finitepast := X_{-M}\dots X_{-1})$ and future contexts $(\finitefuture := X_{0}\dots X_{M-1})$ of some finite length $M$.\footnote{It is not crucial that past and future contexts have the same lengths, and indeed \citet{still-optimal-2010} do not assume this. We do assume equal length throughout this paper for simplicity of our experiments, though nothing depends on this.}
The PRD objective (\ref{eq:ib}) then becomes (\ref{eq:ib-oce}), aiming to predict length-$M$ finite futures from summary codes $Z$ of length-$M$ finite pasts:
\begin{equation}\label{eq:ib-oce}
	\max_{Z : Z \bot \finitefuture | \finitepast} \left( \operatorname{I}[Z, \finitefuture] - \lambda \cdot \operatorname{I}[\finitepast, Z] \right).
\end{equation}
Second, OCF estimates information measures directly from the observed counts of $(\finitepast), (\finitefuture)$ using the plug-in estimator of mutual information. %\rljf{Should be \dots instead of commas?}
With such an estimator, the problem in (\ref{eq:ib-oce}) can be solved using a variant of the Blahut-Arimoto algorithm \citep{tishby-information-1999}, obtaining an encoder $P(Z|\finitepast)$ that maps each observed past sequence $\finitepast$ to a distribution over a (finite) set of code words $Z$.

@@

As in OCF (Section~\ref{sec:OCF}), we assume that a set of finite sample trajectories $x_{-M} \dots x_{M-1}$ is available, and we aim to compress pasts of length $M$ to predict futures of length $M$. To carry this out, we restrict the PRD objective~(\ref{eq:ib}) to such finite-length pasts and futures: %\rljf{I think it would be a lot clearer to use the notation from other papers where it's like $\past$ except the arrow points to $M$. It's hard visually looking at the equations to distinguish $-M\dots-1$ from $0\dots M-1$, and that makes it harder to read. $^M\overleftarrow{X}$}
\begin{equation}\label{eq:ib-oce-repeat}
	\max_{Z : Z \bot \finitefuture | \finitepast} \left( \operatorname{I}[Z, \finitefuture] - \lambda \cdot \operatorname{I}[\finitepast, Z] \right).
\end{equation}
%where $Z$ ranges over all random variables such that the Markovcondition~(\ref{eq:markov}) holds:
%\begin{equation}
%    Z\bot \finitefuture | \finitepast
%\end{equation}
It will be convenient to equivalently rewrite (\ref{eq:ib-oce-repeat}) as
\begin{equation}\label{eq:prd-loss}
	\min_{Z : Z \bot \finitefuture | \finitepast}	\left[\operatorname{H}[\finitefuture | Z] + \lambda \cdot\operatorname{I}[ \finitepast, Z]\right],
\end{equation}
where $\operatorname{H}[\finitefuture | Z]$ is the \key{prediction loss}. Note that minimizing prediction loss is equivalent to maximizing predictiveness $\operatorname{I}[\finitefuture | Z]$. % is the reduction in prediction loss when predicting

@@

Let us assume that we already have some encoding distribution
\begin{equation}
Z \sim P_\phi(Z|\finitepast),
\end{equation}
where $\phi$ is the encoder, expressed in some family of function approximators. The encoder transduces an observation sequence $\finitepast$ into the parameters of the distribution $P_\phi(\cdot|\finitepast)$.
From this encoding distribution, one can obtain the optimal decoding distribution over future observations via Bayes' rule: %, using the Markov condition: 
\begin{equation}
	\label{eq:cond-fut}
	\begin{split}
	P(\finitefuture|Z) &= \frac{P(\finitefuture, Z)}{P(Z)} = \frac{\E_{\finitepast} P(\finitefuture, Z|\finitepast)}{\E_{\finitepast} P_\phi(Z|\finitepast)} =^{(*)} \frac{\E_{\finitepast}   P(\finitefuture|\finitepast)     P_\phi(Z|\finitepast)}{\E_{\finitepast} P_\phi(Z|\finitepast)} \\
%	&= \frac{\E_{X_{-M \dots M-1}} P_\phi(Z|\finitepast)}{\E_{\finitepast} P_\phi(Z|\finitepast)}
	\end{split}
\end{equation}
where $(*)$ uses the Markov condition $Z \bot \finitefuture | \finitepast$.
However, neither of the two expectations in the last line of~(\ref{eq:cond-fut}) is tractable, as they require summation over exponentially many sequences, and algorithms (e.g., dynamic programming) to compute this sum efficiently are not available in general.
For a similar reason, the rate $\operatorname{I}[\finitepast, Z]$ of the code $Z \sim P_\phi(Z|\finitepast)$ is also generally intractable.
%If $\phi$ is given by a powerful function approximating family such as recurrent neural networks,

@@

From a solution to (\ref{eq:bound}), one obtains a solution to (\ref{eq:prd-loss}) by setting $Z \sim P_\phi(\cdot|\finitepast)$.
%The second proposition guarantees that, for a solution to  (\ref{eq:prd-loss}), 
The rate of this code is given as follows
\begin{equation}\label{eq:eq1-new}
 \operatorname{I}[Z, \finitepast] =   \mathbb{E}_{\finitepast}\left[ \operatorname{D_{KL}}\infdivx{P_\phi(Z|\finitepast)}{q(Z)}\right]
\end{equation}
and the prediction loss is given by
\begin{equation}\label{eq:eq2-new}
  %H[X_{0\dots M} | Z] =  -	\mathbb{E}_{z \sim \phi(\finitepast)}\left[\log P_\psi(X_{0\dots M} | z)\right]
  \operatorname{H}[\finitefuture|Z] =   - \mathbb{E}_{X_{-M\dots M-1}}	\mathbb{E}_{Z \sim P_\phi(\finitepast)}\left[\log P_\psi(\finitefuture | Z)\right]. %H[\finitefuture|\finitepast] % \operatorname{H}[\finitefuture]
\end{equation}
\end{corollary}

@@

If $\phi, \psi, q$ are not exact solutions to (\ref{eq:bound}), the two propositions guarantee that we still have bounds on rate and prediction loss for the code $Z$ generated by $\phi$:
\begin{equation}\label{eq:bound1-new}
 \operatorname{I}[Z, \finitepast] \leq   \mathbb{E}_{X}\left[ \operatorname{D_{KL}}\infdivx{P_\phi(Z|\finitepast)}{q(Z)}\right]
\end{equation}
\begin{equation}\label{eq:bound2-new}
  %I[X_{-M\dots 0}, X_{0\dots M} | Z] \leq -	\mathbb{E}_{z \sim \phi(\finitepast)}\left[\log P_\psi(X_{0\dots M} | z)\right]- H[\finitefuture|\finitepast]
  \operatorname{H}[\finitefuture|Z] \leq   - \mathbb{E}_{X_{-M\dots M-1}}	\mathbb{E}_{Z \sim P_\phi(\finitepast)}\left[\log P_\psi(\finitefuture | Z)\right]. % \operatorname{H}[\finitefuture]
\end{equation}
To carry out the optimization~(\ref{eq:bound}), we will restrict $\phi, \psi, q$ to a powerful family of parametric families of function approximators, within which we will optimize the objective with gradient descent.
While the solutions may not be exact solutions to the nonparametric objective~(\ref{eq:bound}), they will still satisfy the bounds (\ref{eq:bound1-new}-\ref{eq:bound2-new}), and---if the family of approximators is sufficiently rich---can come close to turning these into the equalities (\ref{eq:eq1-new}-\ref{eq:eq2-new}).

@@

Given the estimate for prediction loss, we estimate predictiveness $\operatorname{I}[Z, \finitefuture]$ with the following method.
%by estimating $\operatorname{H}[\finitefuture]$ 
Together with $\psi$, we also create an LSTM network that estimates the marginal probability of future observation sequences $P(\finitefuture)$. Let $P_\eta$ be this estimated distribution; we then estimate the predictiveness $\operatorname{I}[Z, \finitefuture] = \operatorname{H}[\finitefuture] - \operatorname{H}[\finitefuture|Z]$ by the difference between the corresponding cross-entropies on the test set~(\cite{mcallester2018formal}):
%$\operatorname{H}[\finitefuture]$ by the cross-entropy of $P_\eta$ on the test data:
\begin{equation}\label{eq:bound-mc-predictiveness}
- \mathbb{E}_{X_{-M\dotsM-1}}\left[\log P_\eta(\finitefuture) - P_\psi(\finitefuture | Z)\right]
\end{equation}
which we approximate using Monte-Carlo sampling on the test set as in~(\ref{eq:bound-mc-rate}-\ref{eq:bound-mc-distortion}).

@@

We study prediction at the range of the words in individual sentences.
As in the previous experiments, we limit our computations to sequences of length 30, already improving over OCF by an order of magnitude.
One motivation is that, when directly estimating PRD, computational cost has to increase with the length of sequences considered, making the consideration of sequences of hundreds or thousands of words computationally infeasible.
Another motivation for this is that we are ultimately interested in predictive rate-distortion as a model of memory in human processing of grammatical structure, formalizing psycholinguistic models of how humans proess individual sentences \citep{gibson-linguistic-1998,futrell-noisy-context-2017}, and linking to studies of the relation between information theory and grammar \citep{koplenig2017statistical}.

@@

\begin{figure*}
\includegraphics[width=0.3\textwidth]{code/figures/english-info.pdf}
%\includegraphics[width=0.45\textwidth]{code/figures/english-beta-mem.pdf}
\includegraphics[width=0.3\textwidth]{code/figures/english-nlogbeta-mem.pdf}
\includegraphics[width=0.3\textwidth]{code/figures/english-nlogbeta-ee.pdf}
	\caption{Left: Rate-Predictiveness for English POS modeling. Center and right: Rate (center) and Predictiveness (right) on English POS Modeling, as a function of $-\log \lambda$. As $\lambda \rightarrow 0$, NPRD (red, $M=15$) continues to discover structure, while OCF (blue, plotted for $M=1,2,3$) exhausts its capacity. }\label{fig:eng-pos}
\end{figure*}%\rljf{It's 1/beta in the figure captions and 1/\lambda in the text}

@@

The rate-predictiveness curve (left) shows that, at low rates, predictiveness is approximately proportional to the rate.
At greater degrees of predictiveness, rate grows faster and faster, whereas predictiveness seems to asymptote to $\approx 1.1$ nats.
The asymptote of predictiveness can be identified with the mutual information between past and future observations, $E_0 := \operatorname{I}[\finitepast, \finitefuture]$, which is a lower bound on the excess entropy. The rate should asymptote to the statistical complexity.
Judging by the curve, natural language---at the time scale we are measuring in this experiment---has a statistical complexity much higher than its excess entropy:
At the highest rate measured by NPRD in our experiment, rate is about 20 nats, whereas predictiveness is about 1.1 nats.
If these values are correct, then---due to the convexity of the rate-predictivity curve---statistical complexity exceeds the excess entropy by a factor of at least $\frac{20}{1.1}$. %\rljf{On precedents for this finding: there's an oblique reference in Grassberger (1989) that excess entropy is ``a very weak [lower bound on SC] in most cases where it was studied'', but following the citations given there I can't find anything.}
Note that this picture agrees qualitatively with the OCF results, which suggest a lower-bound on the ratio of at least $\frac{2.5}{0.6} > 5$.

@@

Let us consider this more quantitatively.
Based on Figure~\ref{fig:eng-pos} (center), we make the ansatz that the map from $\log\frac{1}{\lambda}$ to the rate $R := \operatorname{I}[\finitepast, Z]$ is superlinear:
%\rljf{Was there a change of variables between $\beta$ and $\lambda$ here?}
\begin{equation}\label{eq:r-alpha-beta}
	R = \alpha \left(\log\frac{1}{\lambda}\right)^\beta,
\end{equation}
	with $\alpha>0, \beta>1$.
We fitted $R \approx \left(\log\frac{1}{\lambda}\right)^{1.7}$ ($\alpha=1$, $\beta=1.7$).
Equivalently
\begin{equation}
\frac{1}{\lambda} = \exp\left(\frac{1}{\alpha^{1/\beta}} R^{1/\beta}\right).
\end{equation}
From this, we can derive expressions for rate $R := \operatorname{I}[\finitepast, Z]$ and predictiveness $P := \operatorname{I}[Z, \finitefuture]$ as follows.
For the solution of Predictive Rate-Distortion~(\ref{eq:ib-oce-repeat}), we have
\begin{equation}
	\frac{\partial P}{\partial \theta} - \lambda \frac{\partial R}{\partial \theta} = 0,
\end{equation}
where $\theta$ is the codebook defining the encoding distribution $P(Z|\finitepast)$, and thus
\begin{equation}
\lambda =	\frac{\partial P}{\partial R}.
\end{equation}
Our ansatz therefore leads to the equation
\begin{equation}\label{eq:derivatives}
\frac{\partial P}{\partial R} = \exp\left(-\frac{1}{\alpha^{1/\beta}} R^{1/\beta}\right). %&\ \ \ \partial R/\partial P = \exp\left(\frac{1}{\alpha^{1/\beta}} R^{1/\beta}\right)
\end{equation}
Qualitatively, this says that predictiveness $P$ asymptotes to a finite value, whereas rate $R$---which should asymptote to the statistical complexity---is unbounded.

@@

As the window size $M \rightarrow \infty$, (\ref{eq:sol}) behaves as
\begin{equation}
	P \approx E_0 - \alpha\beta \cdot (R/\alpha)^{\frac{\beta-1}{\beta}} e^{-(R/\alpha)^{1/\beta}},
\end{equation}
which asymptotes to $E_0$.
Given the values fitted above ($\alpha=1$, $\beta=1.7$), we found that $E_0 = 1.13$ yielded a good fit.
Using~(\ref{eq:r-alpha-beta}), this can be extended without further parameters to the third curve in Figure~\ref{fig:eng-pos}.
Resulting fits are shown in Figure~\ref{fig:eng-pos-fitted}.
%$E = E_0 - \Gamma(1.6, (\log\beta))$

@@

Why might language show such behavior?
One natural idea is that the causal states of natural language are the singletons $\past$; that is, no two past trajectories induce a fully equivalent distribution over future trajectories.
Then the code $Z^{(T)}$ which stores the block $X_{-T...-1}$ has rate $R_T = \Theta(T)$, but its predictiveness $P_T =	\operatorname{I}[X_{-T...-1}, X_{0,...M-1}]$ asymptotes to $E_0$.

%By writing $I_t := \operatorname{I}[X_0, X_t | X_1 ... X_{t-1}]$, and repeatedly applying the chain rule of conditional mutual information, we obtain
%\begin{equation}
%	P = \sum_{t_- = -T}^{-1} \sum_{t_+ = 0}^\infty \operatorname{I}[X_{t_+},X_{t_-}|X_{t_-+1}\dots X_{t_+-1}] = \sum_{t_- = -T}^{-1} \sum_{t_+ = 0}^\infty I_{t_+-t_-}  = \sum_{t=1}^\infty \min(t,T) I_t = E_0 - \sum_{t > T} (t-T) I_t
%\end{equation}
%where $E_0$ is again the excess entropy.
%Again, this asymptotes to the $E_0$, but never achieves predictiveness $E_0$ at finite $T$.
%The code has rate
%\begin{equation}
%	R =	\operatorname{H}[X_{-T...-1}]
%\end{equation}
%which is lower bounded by $T$ times the entropy rate: $R \geq T \cdot \operatorname{H}[X_0|\past]$, and thus is unbounded, assuming that the entropy rate is nonzero.
%As $T$ grows, this code thus shows diminishing returns in predictiveness $P$ at linear growth in rate $R$.
%TODO \rljf{A few cites on previous work arguing about whether natural language has nonzero entropy rate. A conference paper by Ryosuke Takahira has a lit review}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\authorcontributions{M.H. and R.F. conceived of and designed the experiments. M.H. performed the experiments and analysed data. M.H. and R.F. wrote the paper. All authors have read and approved the final manuscript.}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% optional
\sampleavailability{Code is available.}

@@

%# also consider
%# https://papers.nips.cc/paper/7210-neural-discrete-representation-learning.pdf

>>>>

%\newcounter{theorem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{question}[theorem]{Question}
\newtheorem{example}[theorem]{Example}
\newtheorem{defin}[theorem]{Definition}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{thm}[theorem]{Theorem}

@@

@@

The distortion is the loss in predictive accuracy when predicting from $Z$, relative to optimal prediction from the full past $\past$.
In the Information Bottleneck formalization, this is equivalent to the amount of mutual information between past and future that is \emph{not} captured by $Z$ \citep{harremoes2007information}: 
\begin{equation}
	\operatorname{I}[\past, \future|Z].
\end{equation}
Due to the Markov condition, the distortion measure satisfies the relation
\begin{equation}
	\operatorname{I}[\past, \future|Z] = \operatorname{I}[\past, \future] - \operatorname{I}[Z, \future],
\end{equation}
i.e., it captures how much less information $Z$ carries about the future $\future$ compared to the full past $\past$.
For a fixed process $(X_t)_t$, choosing $Z$ to minimize the distortion is equivalent to maximizing the mutual information between the code and the future:
\begin{equation}\label{eq:predictiveness}
\operatorname{I}[Z, \future]
\end{equation}
We will refer to~(\ref{eq:predictiveness}) as the \key{predictiveness} of the code $Z$.

@@

As discussed in Section~\ref{sec:complexity}, estimating predictive rate-distortion is related to the problem of estimating statistical complexity.
\citet{clarke2003application} and \citet{still-optimal-2010} consider the problem of estimating statistical complexity from finite data.
While statistical complexity is hard to identify from finite data in general, \citet{clarke2003application} introduce certain restrictions on the underlying process that make this more feasible.

@@

The main prior method for estimating predictive rate-distortion from data is Optimal Causal Filtering (OCF, \citet{still-optimal-2010}).
This method approximates predictive rate-distortion using two approximations:
First, it replaces semi-infinite pasts and futures with bounded-length contexts, i.e., pairs of finite past contexts $(\finitepast := X_{-M}\dots X_{-1})$ and future contexts $(\finitefuture := X_{0}\dots X_{M-1})$ of some finite length $M$.\footnote{It is not crucial that past and future contexts have the same lengths, and indeed \citet{still-optimal-2010} do not assume this. We do assume equal length throughout this paper for simplicity of our experiments, though nothing depends on this.}
The PRD objective (\ref{eq:ib}) then becomes (\ref{eq:ib-oce}), aiming to predict length-$M$ finite futures from summary codes $Z$ of length-$M$ finite pasts:
\begin{equation}\label{eq:ib-oce}
	\max_{Z : Z \bot \finitefuture | \finitepast} \left( \operatorname{I}[Z, \finitefuture] - \lambda \cdot \operatorname{I}[\finitepast, Z] \right).
\end{equation}
Second, OCF estimates information measures directly from the observed counts of $(\finitepast), (\finitefuture)$ using the plug-in estimator of mutual information. 
With such an estimator, the problem in (\ref{eq:ib-oce}) can be solved using a variant of the Blahut-Arimoto algorithm \citep{tishby-information-1999}, obtaining an encoder $P(Z|\finitepast)$ that maps each observed past sequence $\finitepast$ to a distribution over a (finite) set of code words $Z$.

@@

As in OCF (Section~\ref{sec:OCF}), we assume that a set of finite sample trajectories $x_{-M} \dots x_{M-1}$ is available, and we aim to compress pasts of length $M$ to predict futures of length $M$. To carry this out, we restrict the PRD objective~(\ref{eq:ib}) to such finite-length pasts and futures: 
\begin{equation}\label{eq:ib-oce-repeat}
	\max_{Z : Z \bot \finitefuture | \finitepast} \left( \operatorname{I}[Z, \finitefuture] - \lambda \cdot \operatorname{I}[\finitepast, Z] \right).
\end{equation}
%where $Z$ ranges over all random variables such that the Markovcondition~(\ref{eq:markov}) holds:
%\begin{equation}
%    Z\bot \finitefuture | \finitepast
%\end{equation}
It will be convenient to equivalently rewrite (\ref{eq:ib-oce-repeat}) as
\begin{equation}\label{eq:prd-loss}
	\min_{Z : Z \bot \finitefuture | \finitepast}	\left[\operatorname{H}[\finitefuture | Z] + \lambda \cdot\operatorname{I}[ \finitepast, Z]\right],
\end{equation}
where $\operatorname{H}[\finitefuture | Z]$ is the \key{prediction loss}. Note that minimizing prediction loss is equivalent to maximizing predictiveness $\operatorname{I}[\finitefuture | Z]$. % is the reduction in prediction loss when predicting

@@

Let us assume that we already have some encoding distribution
\begin{equation}
Z \sim P_\phi(Z|\finitepast),
\end{equation}
where $\phi$ is the encoder, expressed in some family of function approximators. The encoder transduces an observation sequence $\finitepast$ into the parameters of the distribution $P_\phi(\cdot|\finitepast)$.
From this encoding distribution, one can obtain the optimal decoding distribution over future observations via Bayes' rule: %, using the Markov condition: 
\begin{equation}
	\label{eq:cond-fut}
	\begin{split}
	P(\finitefuture|Z) &= \frac{P(\finitefuture, Z)}{P(Z)} = \frac{\E_{\finitepast} P(\finitefuture, Z|\finitepast)}{\E_{\finitepast} P_\phi(Z|\finitepast)} =^{(*)} \frac{\E_{\finitepast}   P(\finitefuture|\finitepast)     P_\phi(Z|\finitepast)}{\E_{\finitepast} P_\phi(Z|\finitepast)} \\
%	&= \frac{\E_{X_{-M \dots M-1}} P_\phi(Z|\finitepast)}{\E_{\finitepast} P_\phi(Z|\finitepast)}
	\end{split}
\end{equation}
where $(*)$ uses the Markov condition $Z \bot \finitefuture | \finitepast$.
However, neither of the two expectations in the last expression of~(\ref{eq:cond-fut}) is tractable, as they require summation over exponentially many sequences, and algorithms (e.g., dynamic programming) to compute this sum efficiently are not available in general.
For a similar reason, the rate $\operatorname{I}[\finitepast, Z]$ of the code $Z \sim P_\phi(Z|\finitepast)$ is also generally intractable.
%If $\phi$ is given by a powerful function approximating family such as recurrent neural networks,

@@

From a solution to (\ref{eq:bound}), one obtains a solution to (\ref{eq:prd-loss}) by setting $Z \sim P_\phi(\cdot|\finitepast)$.
%The second proposition guarantees that, for a solution to  (\ref{eq:prd-loss}), 
The rate of this code is given as follows
\begin{equation}\label{eq:eq1-new}
 \operatorname{I}[Z, \finitepast] =   \mathbb{E}_{\finitepast}\left[ \operatorname{D_{KL}}\infdivx{P_\phi(Z|\finitepast)}{q(Z)}\right]
\end{equation}
and the prediction loss is given by
\begin{equation}\label{eq:eq2-new}
  %H[X_{0\dots M} | Z] =  -	\mathbb{E}_{z \sim \phi(\finitepast)}\left[\log P_\psi(X_{0\dots M} | z)\right]
  \operatorname{H}[\finitefuture|Z] =   - \mathbb{E}_{X_{\finitepast,\finitefuture}}	\mathbb{E}_{Z \sim P_\phi(\finitepast)}\left[\log P_\psi(\finitefuture | Z)\right]. %H[\finitefuture|\finitepast] % \operatorname{H}[\finitefuture]
\end{equation}
\end{corollary}

@@

If $\phi, \psi, q$ are not exact solutions to (\ref{eq:bound}), the two propositions guarantee that we still have bounds on rate and prediction loss for the code $Z$ generated by $\phi$:
\begin{equation}\label{eq:bound1-new}
 \operatorname{I}[Z, \finitepast] \leq   \mathbb{E}_{X}\left[ \operatorname{D_{KL}}\infdivx{P_\phi(Z|\finitepast)}{q(Z)}\right]
\end{equation}
\begin{equation}\label{eq:bound2-new}
  %I[X_{-M\dots 0}, X_{0\dots M} | Z] \leq -	\mathbb{E}_{z \sim \phi(\finitepast)}\left[\log P_\psi(X_{0\dots M} | z)\right]- H[\finitefuture|\finitepast]
  \operatorname{H}[\finitefuture|Z] \leq   - \mathbb{E}_{X_{\finitepast, \finitefuture}}	\mathbb{E}_{Z \sim P_\phi(\finitepast)}\left[\log P_\psi(\finitefuture | Z)\right]. % \operatorname{H}[\finitefuture]
\end{equation}
To carry out the optimization~(\ref{eq:bound}), we will restrict $\phi, \psi, q$ to a powerful family of parametric families of function approximators, within which we will optimize the objective with gradient descent.
While the solutions may not be exact solutions to the nonparametric objective~(\ref{eq:bound}), they will still satisfy the bounds (\ref{eq:bound1-new}-\ref{eq:bound2-new}), and---if the family of approximators is sufficiently rich---can come close to turning these into the equalities (\ref{eq:eq1-new}-\ref{eq:eq2-new}).

@@

Given the estimate for prediction loss, we estimate predictiveness $\operatorname{I}[Z, \finitefuture]$ with the following method.
%by estimating $\operatorname{H}[\finitefuture]$ 
Together with $\psi$, we also create an LSTM network that estimates the marginal probability of future observation sequences $P(\finitefuture)$. Let $P_\eta$ be this estimated distribution; we then estimate the predictiveness $\operatorname{I}[Z, \finitefuture] = \operatorname{H}[\finitefuture] - \operatorname{H}[\finitefuture|Z]$ by the difference between the corresponding cross-entropies on the test set~(\cite{mcallester2018formal}):
%$\operatorname{H}[\finitefuture]$ by the cross-entropy of $P_\eta$ on the test data:
\begin{equation}\label{eq:bound-mc-predictiveness}
- \mathbb{E}_{X_{\finitepast\finitefuture}}\left[\log P_\eta(\finitefuture) - \log P_\psi(\finitefuture | Z)\right]
\end{equation}
which we approximate using Monte-Carlo sampling on the test set as in~(\ref{eq:bound-mc-rate}-\ref{eq:bound-mc-distortion}).

@@

We study prediction at the range of the words in individual sentences.
As in the previous experiments, we limit our computations to sequences of length 30, already improving over OCF by an order of magnitude.
One motivation is that, when directly estimating PRD, computational cost has to increase with the length of sequences considered, making the consideration of sequences of hundreds or thousands of words computationally infeasible.
Another motivation for this is that we are ultimately interested in predictive rate-distortion as a model of memory in human processing of grammatical structure, formalizing psycholinguistic models of how humans process individual sentences \citep{gibson-linguistic-1998,futrell-noisy-context-2017}, and linking to studies of the relation between information theory and grammar \citep{koplenig2017statistical}.

@@

\begin{figure*}
\includegraphics[width=0.3\textwidth]{code/figures/english-info.pdf}
%\includegraphics[width=0.45\textwidth]{code/figures/english-beta-mem.pdf}
\includegraphics[width=0.3\textwidth]{code/figures/english-nlogbeta-mem.pdf}
\includegraphics[width=0.3\textwidth]{code/figures/english-nlogbeta-ee.pdf}
	\caption{Left: Rate-Predictiveness for English POS modeling. Center and right: Rate (center) and Predictiveness (right) on English POS Modeling, as a function of $-\log \lambda$. As $\lambda \rightarrow 0$, NPRD (red, $M=15$) continues to discover structure, while OCF (blue, plotted for $M=1,2,3$) exhausts its capacity. }\label{fig:eng-pos}
\end{figure*}

@@

The rate-predictiveness curve (left) shows that, at low rates, predictiveness is approximately proportional to the rate.
At greater degrees of predictiveness, rate grows faster and faster, whereas predictiveness seems to asymptote to $\approx 1.1$ nats.
The asymptote of predictiveness can be identified with the mutual information between past and future observations, $E_0 := \operatorname{I}[\finitepast, \finitefuture]$, which is a lower bound on the excess entropy. The rate should asymptote to the statistical complexity.
Judging by the curve, natural language---at the time scale we are measuring in this experiment---has a statistical complexity much higher than its excess entropy:
At the highest rate measured by NPRD in our experiment, rate is about 20 nats, whereas predictiveness is about 1.1 nats.
If these values are correct, then---due to the convexity of the rate-predictivity curve---statistical complexity exceeds the excess entropy by a factor of at least $\frac{20}{1.1}$. 
Note that this picture agrees qualitatively with the OCF results, which suggest a lower-bound on the ratio of at least $\frac{2.5}{0.6} > 5$.

@@

Let us consider this more quantitatively.
Based on Figure~\ref{fig:eng-pos} (center), we make the ansatz that the map from $\log\frac{1}{\lambda}$ to the rate $R := \operatorname{I}[\finitepast, Z]$ is superlinear:
\begin{equation}\label{eq:r-alpha-beta}
	R = \alpha \left(\log\frac{1}{\lambda}\right)^\beta,
\end{equation}
	with $\alpha>0, \beta>1$.
We fitted $R \approx \left(\log\frac{1}{\lambda}\right)^{1.7}$ ($\alpha=1$, $\beta=1.7$).
Equivalently
\begin{equation}
\frac{1}{\lambda} = \exp\left(\frac{1}{\alpha^{1/\beta}} R^{1/\beta}\right).
\end{equation}
From this, we can derive expressions for rate $R := \operatorname{I}[\finitepast, Z]$ and predictiveness $P := \operatorname{I}[Z, \finitefuture]$ as follows.
For the solution of Predictive Rate-Distortion~(\ref{eq:ib-oce-repeat}), we have
\begin{equation}
	\frac{\partial P}{\partial \theta} - \lambda \frac{\partial R}{\partial \theta} = 0,
\end{equation}
where $\theta$ is the codebook defining the encoding distribution $P(Z|\finitepast)$, and thus
\begin{equation}
\lambda =	\frac{\partial P}{\partial R}.
\end{equation}
Our ansatz therefore leads to the equation
\begin{equation}\label{eq:derivatives}
\frac{\partial P}{\partial R} = \exp\left(-\frac{1}{\alpha^{1/\beta}} R^{1/\beta}\right). %&\ \ \ \partial R/\partial P = \exp\left(\frac{1}{\alpha^{1/\beta}} R^{1/\beta}\right)
\end{equation}
Qualitatively, this says that predictiveness $P$ asymptotes to a finite value, whereas rate $R$---which should asymptote to the statistical complexity---is unbounded.

@@

%As the window size $M \rightarrow \infty$, (\ref{eq:sol}) behaves as
%\begin{equation}
%	P \approx E_0 - \alpha\beta \cdot (R/\alpha)^{\frac{\beta-1}{\beta}} e^{-(R/\alpha)^{1/\beta}},
%\end{equation}
%which asymptotes to $E_0$.
Given the values fitted above ($\alpha=1$, $\beta=1.7$), we found that $E_0 = 1.13$ yielded a good fit.
Using~(\ref{eq:r-alpha-beta}), this can be extended without further parameters to the third curve in Figure~\ref{fig:eng-pos}.
Resulting fits are shown in Figure~\ref{fig:eng-pos-fitted}.
%$E = E_0 - \Gamma(1.6, (\log\beta))$

@@

%Why might language show such behavior?
%One natural idea is that the causal states of natural language are the singletons $\past$; that is, no two past trajectories induce a fully equivalent distribution over future trajectories.
%Then the code $Z^{(T)}$ which stores the block $X_{-T...-1}$ has rate $R_T = \Theta(T)$, but its predictiveness $P_T =	\operatorname{I}[X_{-T...-1}, X_{0,...M-1}]$ asymptotes to $E_0$.

%By writing $I_t := \operatorname{I}[X_0, X_t | X_1 ... X_{t-1}]$, and repeatedly applying the chain rule of conditional mutual information, we obtain
%\begin{equation}
%	P = \sum_{t_- = -T}^{-1} \sum_{t_+ = 0}^\infty \operatorname{I}[X_{t_+},X_{t_-}|X_{t_-+1}\dots X_{t_+-1}] = \sum_{t_- = -T}^{-1} \sum_{t_+ = 0}^\infty I_{t_+-t_-}  = \sum_{t=1}^\infty \min(t,T) I_t = E_0 - \sum_{t > T} (t-T) I_t
%\end{equation}
%where $E_0$ is again the excess entropy.
%Again, this asymptotes to the $E_0$, but never achieves predictiveness $E_0$ at finite $T$.
%The code has rate
%\begin{equation}
%	R =	\operatorname{H}[X_{-T...-1}]
%\end{equation}
%which is lower bounded by $T$ times the entropy rate: $R \geq T \cdot \operatorname{H}[X_0|\past]$, and thus is unbounded, assuming that the entropy rate is nonzero.
%As $T$ grows, this code thus shows diminishing returns in predictiveness $P$ at linear growth in rate $R$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\authorcontributions{Methodology, Michael Hahn; Writing – original draft, Michael Hahn and Richard Futrell.}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% optional
%\sampleavailability{Code is available.}

@@
